Future challenges for Research Software Engineering

### 2025-05-12 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

The digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\begin{FramedParagraphWithFootnotes}
GC-NOTE: With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernization, cleanup, and modularization,
which can pay off in faster and larger development in the future.
This is not possible by short-term funded solo RSEs.
\end{FramedParagraphWithFootnotes}

\subsection{Education, and changes in the educational system}

\begin{FramedParagraphWithFootnotes}
FG-NOTE: I wonder if we need a separation between education of RSEs and changes in the education of the broader society.
\end{FramedParagraphWithFootnotes}

\begin{FramedParagraphWithFootnotes}
GC-NOTE: Digital competencies in the general population continue to increase, making jumping to software engineering easier than before.

GC-NOTE: By that point, we have an RSE Master's.
\end{FramedParagraphWithFootnotes}

\subsection{Impact of AI and automation}
Is it better than cheap PhDs?

\begin{FramedParagraphWithFootnotes}
FG-NOTE: AI will change the interaction with researchers at a fundamental level. Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
I foresee that AI drives this low-key interaction to a highly specialized level of knowledge, since the availability of AI is far better.
\end{FramedParagraphWithFootnotes}

\begin{FramedParagraphWithFootnotes}
GC-NOTE: As we continue and expand documenting our work in the open and making our code available,
there will be more structured material to train GenAI on.

However, as people get quicker answers from GenAI, the community resources (Stack Overflow, Wikipedia) get poorer.
\end{FramedParagraphWithFootnotes}

@@

MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts? (GC: Where "product/artifacts", maybe also "publications"?)

@@

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay have started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is working in a project that develops
software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
at her university, and increasingly struggles to find compelling Open Source alternatives that lets them keep their data
to software that is backed by large corporate entities. So far she still has the support of her university board, since it values
strategic autonomy.
Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master's courses,
but quickly decided that work in academia with its time limited contracts is not attractive.
Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
and therefore have deprioritized publicly funded research. Kyle decided to go to a defense company since they offer a
more secure employment in this polarized world.
\end{story}

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{What do we want to have achieved}{academic}
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
At universities which have them, we want to see that researchers naturally accept them as part of their used services.
\end{whatis}

@@

What are open questions here?
Organizations will have to come up with models for hiring people. While
a standardized RSE Master's makes it easier, it will be interesting to see if
RSE are still often hired by word of mouth or if the master's really makes it more comparable.

\begin{FramedParagraphWithFootnotes}
  GC-NOTE: New section on "New challenges":

  \begin{itemize}
    \item Networking of embedded RSEs and RSE units
    \item Everyone needs infrastructure: does everyone need to host their own infrastructure?
    \item Everyone hosts their infrastructure, but then everyone stays in silos? We need federated infrastructure.
  \end{itemize}
\end{FramedParagraphWithFootnotes}

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}

RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units. (TODO: Cite RSE groups paper)
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding is more difficult to acquire in lower levels of an organization.
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, Göttingen).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups themselves prove themselves valuable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

@@

The workshop participants found that in earlier learning stages,
details may be intentionally hidden,
fearing a diminishing capability to handle technical complexity in research environments.
As a result, RSEs invest more effort in making individual research software components
less complex with better user experience.
This made participants worry about growing mismatch of peoples' expectations.
Participants expected an over-reliance on generative AI tools as an extreme manifestation of this issue.
No concrete examples were given on which complexities are now hidden behind abstractions.

We acknowledge but would also like to challenge the observations of the participants.
We believe that ``we are standing on the shoulders of giants'':
By reducing the complexity of individual software components,
we are now able to develop systems of complexity and capability not previously possible,
allowing the scientific community to address questions not previously in grasp.
When it comes to teaching, it is true that students need to learn how to handle complexity,
but it is also important to learn how to abstract from details
by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.
It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
as well as RSEs that can manage complex systems of almost black-box components,
and the future of RSE education should aim to prepare students for both roles.

@@

GC-NOTE: Shall we add another story box here?

@@

GC-NOTE: RSE develop together with and for scientists.
Better software leads to better research,
better research leads to more better software.

>>>>

The digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\subsection{Education, and changes in the educational system}
FG-NOTE: I wonder if we need a separation between education of RSEs and changes in the education of the broader society.

\subsection{Impact of AI and automation}
Is it better than cheap PhDs?
FG-NOTE: AI will change the interaction with researchers at a fundamental level. Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
I foresee that AI drives this low-key interaction to a highly specialized level of knowledge, since the availability of AI is far better.

@@

MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts?

@@

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay have started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is working in a project that develops
software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
at her university, and increasingly struggles to find compelling Open Source alternatives that let's them keep their data
to software that is backed by large corporate entities. So far she still has the support of her university board, since it values
strategic autonomy.
Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master courses,
but quickly decided that work in academia with its time limited contracts is not attractive.
Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
and therefore have deprioritized publicly funded research. Kyle decided to go to a defence company since they offer a
more secure employment in this polarized world.
\end{story}

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{What do we want to have achieved}{academic}
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
At universities which have them, we want to see, that researchers naturally accept them as part of their used services.
\end{whatis}

@@

What are open questions here?
Organizations will have to come up with models for hiring people. While
a standardized RSE masters makes it easier, it will be interesting to see, if
RSE are still often hired by word of mouth or if the master's really makes it more comparable.

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
The RSE is probably one of the first professionalizations in the academic mid-level that is able
to collectively voice their concerns. This enables the possibility to reshape universities.
If more people are employed as mid-level, this makes a steadier pool of people
available that can integrate new people that are regularly coming into departments and know the local facilities.
On the other hand, an upcoming challenge how universities decide to design the contracts of people employed
in RSE units. Will they be permanent, or will there also be an up-or-out mentality?

As outlined in [cite RSE group paper] not all universities need to opt for strengthening
the academic mid-level. Some might have chosen to set up external companies to outsource typical RSE tasks,
thereby providing career paths outside of academia.

FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.

@@

The workshop participants found that in earlier learning stages,
details of software frameworks and libraries may be intentionally hidden,
fearing a diminishing capability to handle technical complexity in research environments.
As a result, RSEs invest more effort in making individual research software components
less complex with better user experience.
This made participants worry about growing mismatch of peoples' expectations.
Participants expected an over-reliance on generative AI tools as an extreme manifestation of this issue.
No concrete examples were given on which complexities are now hidden behind abstractions.

We acknowledge but would also like to challenge the observations of the participants.
We believe that ``we are standing on the shoulders of giants'':
By reducing the complexity of individual software components,
we are now able to develop systems of complexity and capability not previously possible,
allowing the scientific community to address questions not previously in grasp.
When it comes to teaching, it is true that students need to learn how to handle complexity,
but it is also important to learn how to abstract from details,
by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.
It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
as well as RSEs that can manage complex systems of almost black-box components,
and the future of RSE education should aim to prepare students for both roles.

@@

@@
